DNNNNR0: How to Avoid OOTA Without Really Trying

### 2025-05-10 paper/ootavacuous.tex
ootavacuous: Add a section giving examples exercising the definition
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

As long as \co{z} is zero, changes in the
value of \co{y} will not cause a change in the value stored to \co{x}.
As a result, the semantic dependency from \co{y} to \co{x} exists only
in executions where \co{z} is nonzero,
which shows it is a property of the execution, not just of the source code.

@@

>>>>

Expanding on earlier discussions~\cite{Boehm:2014:OGA:2618128.2618134},
as long as \co{z} is zero, changes in the
value of \co{y} will not cause a change in the value stored to \co{x}.
As a result, the semantic dependency from \co{y} to \co{x} exists only
in executions where \co{z} is nonzero,
which shows it is a property of the execution, not just of the source code.

@@

\subsection{Exercising the Definition}
\label{sec:Exercising the Definition}

This section exercises the definition of ``semantic dependency''
presented in the previous section on selected examples.
These exercises use the following steps:

\begin{enumerate}
\item	Identify the potential semantic dependencies of interest.
\item	Compile the code to assembly language, either using a
	real compiler or conceptually.
\item	Check for an assembly level syntactic dependency between
	the loads and stores making of the potential semantic
	dependency.
	\begin{enumerate}
	\item	If there is no assembly level syntactic dependency,
		then there can be no semantic
		dependency.\label{item:No assembly level syntactic dependency}
	\item	Otherwise, if there is no execution having a witness,
		then there can be no semantic
		dependency.\label{item:No witness}
	\item	Otherwise, there is a semantic
		dependency.\label{item:No sdep}
	\end{enumerate}
\end{enumerate}

In cases~\ref{item:No assembly level syntactic dependency}
and~\ref{item:No witness}
there is no semantic dependency.
In case~\ref{item:No sdep},
there might or might not be a semantic dependency, but
time will definitely be consumed executing the resulting code.
Either way, the candidate dependency cannot be part of an OOTA
cycle.

The following sections apply these steps to two of the examples in
this paper.
% Should we call out the relationship to the perturbation method of P0422r0?

\subsubsection{Exercising Function Calls}
\label{sec:Exercising Function Calls}

Section~\ref{sec:Semantic Dependencies and Function Calls}
provides the following example:
\begin{quote}
\begin{verbatim}
y = f(x);
\end{verbatim}
\end{quote}
If the compiler knows nothing about the function \co{f()} and the variable
\co{y}, it must emit code that loads from \co{x}, passes this value to
\co{f()}, and finally stores the return value to \co{y}.
Is there a semantic dependency from the load of \co{x} to the store of
\co{y}?
To determine this, it is necessary to apply this same analysis to the
definition of \co{f()}, treating its argument as a load and each of its
\co{return} statements as a store.
If and only if this analysis shows that, for a given execution, there
is a semantic dependency within \co{f()}, then there must also be a
semantic dependency from the load of \co{x} to the store of \co{y}.
Of course, if the compiler cannot prove that there is no such semantic
dependency, it must assume than there is one.

If \co{f} is a pointer to a function, then the compiler must also
generate load from \co{f}, so that there might be a semantic dependency
from either or both of \co{f} and \co{y} to the store to \co{y}.
This is straightforward:
A given execution will load a pointer to a particular function from
\co{f}, so if there is another execution that loads a pointer to a
different function, and the two functions return different values for
that execution's value from \co{y}, then there is a semantic dependency
from \co{f} to \co{y}.

Situations where the compiler knows about \co{f} more often result in
inlining, in which case there is no function call, allowing the semantic
dependencies (or lack thereof) to be determined with respect to the
inlined code.

\subsubsection{Exercising Multiplication}
\label{sec:Exercising Multiplication}

Sections~\ref{sec:Semantic Dependencies and Source Code}
and~\ref{sec:Semantic Dependencies Can Be Many-To-One}
provide the following example:
\begin{quote}
\begin{verbatim}
x = y * z;
\end{verbatim}
\end{quote}
This has potential semantic dependencies between the loads from \co{x}
and \co{y} to the store to \co{z}.
The corresponding assembly code can vary depending on what the compiler
knows about the values of \co{x} and \co{y}.

If it knows nothing, then it must load from both, multiply the values
loaded, and then store the resulting product to \co{z}.
There is an assembly-level syntactic dependency, and if we start
with the execution in which the value 1 is loaded from both variables
(resulting in the value 1 being stored to \co{z}),
then the execution in which the value 2 is loaded from \co{x} witnesses
a semantic dependency from \co{x} to \co{z}.
Similarly, the execution in which the value 2 is loaded from \co{y}
witnesses a semantic dependency from \co{y} to \co{z}.
Both witnesses result in the value 2 being stored to \co{z}.

But suppose we instead start with the execution where the value 0
is loaded from \co{x} and the value 1 is loaded from \co{y}.
In this case, there is no witness execution that varies only the value
of \co{y}, as all such executions will result in the value 0 being stored
to \co{z}.
However, the execution where the value -1 is loaded from \co{x}
witnesses a semantic dependency from \co{x} to \co{y}, resulting in
the value -1 being stored to \co{z}.
Furthermore, the execution where the value -1 is loaded from \co{x}
and the value -2 is loaded from \co{y} witnesses a semantic dependency
from the set \{\co{x}, \co{y}\} to \co{z}, resulting in the value 2 being
stored to \co{z}.

But suppose we instead start with the execution where the value 0
is loaded from both \co{x} and \co{y}.
There is again no witness execution that varies only the value of \co{y},
but there is also no witness execution that varies only the value of \co{x},
all such executions resulting in the value 0 stil being stored to \co{z}.
However, the execution where the value 2 is loaded from both \co{x} and
\co{y} witnesses a semantic dependency from the set \{\co{x}, \co{y}\}
to \co{z}, resulting in the value 4 being stored to \co{z}.

Alternatively, if the compiler knows that the value loaded from \co{x}
is guaranteed to be zero, then it can omit the loads from both \co{x}
and \co{y} (assuming that the value from \co{y} is not otherwise used),
storing the constant 0 to \co{z}.
There can be no semantic dependencies from these now-nonexistent loads
to the store to \co{z}.
And then any execution that witnessed a semantic dependency from
either or both of \co{x} and \co{y} would indicate a bug in the
compiler, which would have been wrong in its assumption of the
guarantee that \co{x} was zero.
